One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning

### 2023-10-31 Raschtree_MH.Rnw
updates
First submission finalized

\todo{I usually use load for Rda files, Marjolein uses readRDS, is there an advantage? use the same in both examples.
Mirka: from what I understand, readRDS is for loading .rds/.rda files with only one r object in it, while load is for .rda files with one or more r object in it. I would opt for .rda, because I think it is more general? Marjolein: I think if we include datasets in supplemental materials, we should have files which are only 1 dataset, not sets of datasets or objects. I think this will make the examples more simple. In practice, users will probably also have only 1 dataset.}

@@

Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function. The resulting tree is depicted in Figure \ref{fig:MHtree1}. The item parameters for the nine culture items are displayed in the terminal nodes of the tree. However, it is hard to interpret the tree structure because such a large number of splits was created. Moreover, it is hard to tell whether all splits are due to substantial amounts of DIF, or whether some of the splits are due to small DIF effects that only became statistically significant due to the large sample size.

@@

@@

For the purification strategy, there are three options: \texttt{none}, \texttt{2step} and \texttt{iterative}. Purification means that items which have already been diagnosed with DIF are taken out of the sum score, which is used as the matching criterion in the final DIF test. It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the references therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}. In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2. When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached. 
The final set of DIF-free items resulting from the purification is also used for aligning the item parameters for each group comparison in the tree plots with color coding, which are displayed below.

@@

The result is presented in Figure~\label{fig:MHtree2}, which reveals that the Rasch tree is much more concise with a lower number of splits. In addition, in each node the number of items classified as A, B, or C is displayed. For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and 4 items are classified in category C (large DIF).

@@

These items are %(relative to the DIF-free items for this group comparison) 
easier or less difficult to answer for participants with female or missing gender.

@@

Alternatively, we can also color the DIF items in the end node profiles according to a later split in the tree:

>>>>

@@

Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function:

@@

The resulting tree is depicted in Figure \ref{fig:MHtree1}. The item parameters for the nine culture items are displayed in the terminal nodes of the tree. However, it is hard to interpret the tree structure because such a large number of splits was created. Moreover, it is hard to tell whether all splits are due to substantial amounts of DIF, or whether some of the splits are due to small DIF effects that only became statistically significant due to the large sample size.

@@

For the purification strategy, there are three options: \texttt{none}, \texttt{2step} and \texttt{iterative}. Purification means that items which have already been diagnosed with DIF are taken out of the sum score, which is used as the matching criterion in the final DIF test. It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the references therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}. In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2. When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached. 
The final set of DIF-free items resulting from the purification will also be used for aligning the item parameters for each group comparison in the tree plots with color coding, which will be discussed and displayed below.

We fit the Raschtree using the Mantel-Haenszel stopping criterion as follows:

@@

The result is presented in Figure~\ref{fig:MHtree2}, which reveals that the Rasch tree is much more concise with a lower number of splits. In addition, in each node the number of items classified as A, B, or C is displayed. For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and 4 items are classified in category C (large DIF).

@@

These items are %(relative to the DIF-free items for this group comparison) 
easier or less difficult to answer for participants with female or missing gender. Alternatively, we can also color the DIF items in the end node profiles according to a later split in the tree:

@@

### 2023-10-31 _MOB_paper.Rnw
updates
First submission finalized

\linespread{1.3}

\title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
\shorttitle{shorttitle}

\authorsnames{Marjolein Fokkema$^1$, Mirka Henninger$^2$ \& Carolin Strobl$^2$}
\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory (IRT) models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF). %DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness. % Test fairness is more related to use of test scores in practice than group-level comparisons I would think?
We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

@@

## Create long dataset
pred_vars <- "program"
HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
                  varying = list(paste0("PPVT_Raw", 2*(43:52)),
                                 paste0("AgeTest_Yr", 2*(43:52))),
                  direction = "long", 
                  v.names = c("PPVT_Raw", "AgeTest_Yr"))
HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
                   "ChildID", "MotherID", "Program", "PPVT", "Age")
HS_dat <- HS_dat[complete.cases(HS_dat), ]
HS_dat$Age_orig <- HS_dat$Age
#HS_dat$Age <- sqrt(HS_dat$Age)
rownames(HS_dat) <- NULL
saveRDS(HS_dat, file = "HS_dat.Rda")
@

@@

<<>>=
HS_dat <- readRDS("HS_dat.Rda")
head(HS_dat, 3)
@

@@

@@

<<echo=FALSE, results=tex>>=
coefs <- round(coef(HS_tree), digits = 2)
design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
                        Program = factor(rep(c("None", "HS"))))
design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 6), 
                                     re.form = ~0)
design_df$`PPVT at age 12` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 12), 
                                      re.form = ~0)
names(design_df)[1] <- "Node"
knitr::kable(design_df, format = "latex", digits = 2, 
             caption = "Node-specific average predicted PPVT scores at different ages.", 
             label = "predictions", centering = FALSE,
             linesep = "", # linesep command suppressess addlinesep every 5 rows
             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)
@

To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12. All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT. This benefit increases over time for the group with higher mother's AFTQ scores. The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2). For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12. These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14. Fade out is reported to be particularly strong for Afriacn-American and very disadvantaged children.

@@

\newpage
\FloatBarrier
\section{Application Example 2: Subgroup Detection in Rasch Models}
\label{sec:TutorialRasch}

@@

As with any exploratory method, MOB should be considered as a tool for generating hypotheses and not as a confirmatory technique. While the MOB algorithm, as it is implemented in the \texttt{R} packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory hypothesis testing it is recommended to use a new dataset, not used for fitting the tree. Further, resampling or $k$-fold cross validation techniques may be used to assess stability and accuracy of the model. \cite{PhilyRusc18} show how to use these techniques for assessing the stability of the set of selected variables and cutpoints. For GLMM trees, cross-validation can additionally be used for evaluating predictive accuracy \cite[e.g., ]{RooiyWeed20}, or for further optimizing predictive accuracy through the choice of model-fitting parameters, like the minimum number of observations the terminal nodes or the maximum tree depth \cite{JameyWitt23}.

@@

\hyperref[sec:TutorialRasch]{Application Example 2} showed that the Rasch tree method can employ an effect size measure to stop the tree from growing if effect-size differences are non-substantial. This also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large. At the same time, effect sizes are less straightforward to calculate and interpret in mixed-effects models \citep{JuddyWest17}, and thereby also for GLMM trees. \todo{Interpretation methods might not work as expected for multilevel data, so I am not convinced we should discuss this and call it a remedy.}A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots. These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s). A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.

@@

\todo{Add DOIs.}

@@

>>>>

\linespread{1.5}

\title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
\shorttitle{Subgroup Detection Using Model-Based Recursive Partitioning}

\authorsnames{ }
\authorsaffiliations{ }

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity in, for example, intervention studies. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory (IRT) models. In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF). DIF should be investigated before using test results for group comparisons. We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

\begin{titlepage}
   \begin{center}
       \vspace*{1cm}

       \textbf{\large One Model May Not Fit All}\\

       \vspace{0.5cm}
        \large Subgroup Detection Using\\Model-Based Recursive Partitioning\\

       \vspace{1.5cm}

       \textbf{Marjolein Fokkema$^1$, Mirka Henninger$^2$ and Carolin Strobl$^2$}\\
       $^1$Leiden University, $^2$Universit\"at Z\"urich

       \vfill

       \vspace{0.8cm}

   \end{center}
\end{titlepage}

@@

## Create long dataset
pred_vars <- "program"
HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
                  varying = list(paste0("PPVT_Raw", 2*(43:52)),
                                 paste0("AgeTest_Yr", 2*(43:52))),
                  direction = "long", 
                  v.names = c("PPVT_Raw", "AgeTest_Yr"))
HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
                   "ChildID", "MotherID", "Program", "PPVT", "Age")
HS_dat <- HS_dat[complete.cases(HS_dat), ]
HS_dat$Age_orig <- HS_dat$Age
#HS_dat$Age <- sqrt(HS_dat$Age)
rownames(HS_dat) <- NULL
save(HS_dat, file = "HS_dat.Rda")
@

@@

<<>>=
load("HS_dat.Rda")
head(HS_dat, 3)
@

@@

<<echo=FALSE, results=tex>>=
coefs <- round(coef(HS_tree), digits = 2)
design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
                        Program = factor(rep(c("None", "HS"))))
design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 6), 
                                     re.form = ~0)
design_df$`PPVT at age 12` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 12), 
                                      re.form = ~0)
names(design_df)[1] <- "Node"

#library("kableExtra")
#kable_styling(
knitr::kable(design_df, format = "latex", digits = 2, 
             caption = "Node-specific average predicted PPVT scores at different ages.", 
             label = "predictions", centering = FALSE,
             linesep = "", # linesep command suppressess addlinesep every 5 rows
             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)#,
#latex_options = "HOLD_position")
@

@@

To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12. All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT. This benefit increases over time for the group with higher mother's AFTQ scores. The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2). For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12. These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14. Fade out was reported to be particularly strong for Afriacn-American and very disadvantaged children. The global benefit of participation in Head Start as shown in Figure~\ref{fig:global_lmm} and reported by \cite{Demi09} thus seems to differ across groups.

@@

\newpage

\section{Application Example 2: Subgroup Detection in Rasch Models}
\label{sec:TutorialRasch}

@@

As with any exploratory method, MOB should be considered as a tool for generating hypotheses and not as a confirmatory technique. While the MOB algorithm, as it is implemented in the \texttt{R} packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory hypothesis testing it is recommended to use a new dataset, not used for fitting the tree. Further, resampling or $k$-fold cross validation techniques may be used to assess stability and accuracy of the model. \cite{PhilyRusc18} show how to use these techniques for assessing the stability of the set of selected variables and cutpoints. For GLMM trees, cross-validation can additionally be used for evaluating predictive accuracy \citep[e.g., ][]{RooiyWeed20}, or for further optimizing predictive accuracy through the choice of model-fitting parameters, like the minimum number of observations in the terminal nodes or the maximum tree depth \citep{JameyWitt23}.

@@

\hyperref[sec:TutorialRasch]{Application Example 2} showed that the Rasch tree method can employ an effect size measure to stop the tree from growing if effect-size differences are non-substantial. This also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large. For mixed-effects models, computation of effect sizes is far from straightforward and several different approaches have been proposed \citep[e.g., ][]{JuddyWest17, BrysyStev18, Bran18}. Effect-size based stopping criteria for GLMM trees are therefore currently not available, but any of the proposed computation approaches proposed for tradtional GLMMs can also be computed from the parameter estimates of GLMM trees.

%\todo{Interpretation methods might not work as expected for multilevel data, so I am not convinced we should discuss this and call it a remedy.} 
%A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots. These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s). A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.

@@

%\todo{Add DOIs.}

@@

<<echo=FALSE>>=
Stangle(file='_MOB_paper.Rnw', output = '_MOB_paper.R', ,encoding='utf8',
        drop.evalFALSE = TRUE, annotate = FALSE)
@

### 2023-10-31 _MOB_paper.tex
updates
First submission finalized

\begin{Schunk}
\begin{Sinput}
 HS_dat <- readRDS("HS_dat.Rda")
 head(HS_dat, 3)
\end{Sinput}
\begin{Soutput}
       AFTQ     Race   Income Mom_height Mom_edu_yrs ChildID MotherID Program
1  3.478122 Hispanic 37731.07        502          12   20502      205      HS
2  3.478122 Hispanic 37731.07        502          12   20501      205    None
3 15.964368    Black 16119.13        504          10   22403      224    None
  PPVT Age Age_orig
1   18   4        4
2   48   7        7
3   69   7        7
\end{Soutput}
\end{Schunk}

@@

\begin{figure}%
\caption{GLMM tree with trajectories of PPVT scores}
\includegraphics{_MOB_paper-012}

@@

\includegraphics{_MOB_paper-013}
\\\textit{Note}: Red lines depict average node-specific trajectories for siblings participating in Head Start; black lines depict average node-specific trajectories for siblings not participating in Head Start.
\label{fig:lmm_tree}
\end{figure}%

@@

\FloatBarrier

\caption{\label{tab:predictions}Node-specific average predicted PPVT scores at different ages.}
\begin{tabular}[t]{cccc}
\toprule
Node & Program & PPVT at age 6 & PPVT at age 12\\
\midrule
2 & None & 45.70 & 95.63\\
2 & HS & 49.48 & 98.02\\
5 & None & 50.36 & 107.43\\
5 & HS & 53.74 & 104.98\\
6 & None & 53.33 & 107.82\\
6 & HS & 59.99 & 115.70\\
7 & None & 59.64 & 117.13\\
7 & HS & 63.56 & 124.09\\
\bottomrule
\end{tabular}
\end{table}
To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12. All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT. This benefit increases over time for the group with higher mother's AFTQ scores. The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2). For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12. These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14. Fade out is reported to be particularly strong for Afriacn-American and very disadvantaged children.

@@

>>>>

\begin{Schunk}
\begin{Sinput}
 load("HS_dat.Rda")
 head(HS_dat, 3)
\end{Sinput}
\begin{Soutput}
       AFTQ     Race   Income Mom_height Mom_edu_yrs ChildID MotherID Program
1  3.478122 Hispanic 37731.07        502          12   20502      205      HS
2  3.478122 Hispanic 37731.07        502          12   20501      205    None
3 15.964368    Black 16119.13        504          10   22403      224    None
  PPVT Age Age_orig
1   18   4        4
2   48   7        7
3   69   7        7
\end{Soutput}
\end{Schunk}

@@

\caption{\label{tab:predictions}Node-specific average predicted PPVT scores at different ages.}
\begin{tabular}[t]{cccc}
\toprule
Node & Program & PPVT at age 6 & PPVT at age 12\\
\midrule
2 & None & 45.70 & 95.63\\
2 & HS & 49.48 & 98.02\\
5 & None & 50.36 & 107.43\\
5 & HS & 53.74 & 104.98\\
6 & None & 53.33 & 107.82\\
6 & HS & 59.99 & 115.70\\
7 & None & 59.64 & 117.13\\
7 & HS & 63.56 & 124.09\\
\bottomrule
\end{tabular}
\end{table}
\FloatBarrier

\begin{figure}%
\caption{GLMM tree with trajectories of PPVT scores}
\includegraphics{_MOB_paper-013}

@@

\includegraphics{_MOB_paper-014}
\\\textit{Note}: Red lines depict average node-specific trajectories for siblings participating in Head Start; black lines depict average node-specific trajectories for siblings not participating in Head Start.
\label{fig:lmm_tree}
\end{figure}%

@@

To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12. All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT. This benefit increases over time for the group with higher mother's AFTQ scores. The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2). For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12. These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14. Fade out was reported to be particularly strong for Afriacn-American and very disadvantaged children. The global benefit of participation in Head Start as shown in Figure~\ref{fig:global_lmm} and reported by \cite{Demi09} thus seems to differ across groups.

@@

\begin{Schunk}
\begin{Soutput}
Writing to file _MOB_paper.R 
\end{Soutput}
\end{Schunk}
